---
title: "Lab 8"
author: "Alex Cagle"
date: "2023-05-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab 8

```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.data <- read.csv(fna.data, row.names=1)
```

```{r}
head(wisc.data)
```

```{r}
# Q: How many patient samples are there in the dataset?
nrow(wisc.data)
```
There are `r nrow(wisc.data)` patients in the dataset.

```{r}
# Q: How many cancer (M) and non-cancer (B) samples are there?
table(wisc.data$diagnosis)

```
```{r}
diagnosis <- as.factor(wisc.data$diagnosis)
```

```{r}
wisc <- wisc.data[,-1]
```

```{r}
# Q: How many 'dimensions', 'vairables', 'columns' are there in the dataset?
ncol(wisc)
```

## Principal Component Analysis

> PCA aims to take a large set of data and reduce its dimensionality while retaining the relationships present in the data, which can help us analyze data more easily. 

```{r}
round(colMeans(wisc))
```


```{r}
pca <- prcomp(wisc, scale = TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1], pca$x[,2], col=diagnosis)
```
```{r}
# Create a data.frame for ggplot
df <- as.data.frame(pca$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```


> Q: From your results, what proportion of the original variance is captured by the first principal components (PC1)?
> Principal component 1 represents the highest variation. Proportion of variance captured by PC1 = 44.27%

> Q: How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
> 3 dimensions (PC1, PC2, and PC3)

> Q: How many principal components (PCs) are required to describe at least 90% of the original data.
> 7 dimensions (PC1 through PC7)

> Q: How much variance is captured in the top 3 PC's?
> The top 3 PC's capture 72.6% of the variance.


> Q9: For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC.

```{r}
pca$rotation
pca$rotation["concave.points_mean", 1]
```
```{r}
attributes(pca)
```

## Combine PCA results with clustering

We can use our new PCA variables (i.e. the scores along the PCs contained in t `pca$x`) as input for other methods such as clustering.

```{r}
# Hclust needs a distance matrix as input
d <- dist( pca$x[, 1:3])

hc <- hclust(d, method = "ward.D2")
plot(hc)
```

To get our cluster membership vector we can use the `cutree()` function and specify a height (`h`) or number of groups (`k`).

```{r}
grps <- cutree(hc, h=80)
table(grps)
```

I want to find out how many diagnoses "M" and "B" are in each grp?

```{r}
table(diagnosis)
```
```{r}
table(diagnosis, grps)
```
```{r}
plot(pca$x[,1:2], col=grps)
```

```{r}
plot(pca$x[,1:2], col=diagnosis)
```


```{r}
# Create a data.frame for ggplot
df <- as.data.frame(pca$x)

# Make a scatter plot colored by grps
ggplot(df) + 
  aes(PC1, PC2) + 
  geom_point(col=grps)
```

> Q15. What is the specificity and sensitivity of our current results?

```{r}
# group 1 is mostly malignant while group 2 is mostly benign

# sensitivity = TP / (TP + FN)
sensitivity <- 179 / (179 + 24)

# specificity = TN / (TN + FN)
specificity <- 333 / (333 + 24)

print(sensitivity)
print(specificity)
```

## Prediction
```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(pca, newdata=new)
npc
```

```{r}
plot(pca$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q16: Which of these new patients should we prioritize for follow up based on your results?

We should follow up with the data points that are colored in red (i.e. group 1), since they represent the sick patients.