---
title: "Halloween Candy"
author: "Alex Cagle"
date: "2023-05-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q: What are these fruity candy?

```{r}
rownames(candy[candy$fruity == 1, ])
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value? In other words, how often does my favorite candy win?

```{r}
candy["Snickers", ]$winpercent
```
> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
#install.packages('skimr')
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the `winpercent` column is on a 0:100 scale and all others appear to be on a 0:1 scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero here means the candy is not classified as containing chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```


```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=10)
```
> Q9. Is the distribution of winpercent values symmetrical?

No.

> Q10. Is the center of the distribution above or below 50%?

Below 50% with a mean of:

```{r}
mean(candy$winpercent)
```
> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

To answer this, I need to subset the candy dataset to select only chocolate candy and then calculate the mean of these. Then, I need to do the same for fruity candy.

```{r}
# Filter/subset data to just get chocolate rows
chocolate.candy <- candy[as.logical(candy$chocolate), ]

# Get winpercent values
chocolate.winpercent <- chocolate.candy$winpercent

# Calculate the mean of winpercent values
mean(chocolate.winpercent)
```

```{r}
# Filter/subset data to just get fruity rows
fruity.candy <- candy[as.logical(candy$fruity), ]

# Get winpercent values
fruity.winpercent <- fruity.candy$winpercent

# Calculate the mean of winpercent values
mean(fruity.winpercent)
```
So, chocolate (60.92%) is more popular than fruity (44.11%).


> Q12. Is this difference statistically significant?

Yes, the difference is large, so it is probably statistically significant. 


## Overall Candy Ranking

There is a base R function called `sort()` for sorting vectors of input.
```{r}
x <- c(5, 2, 10)

sort(x, decreasing = TRUE)
```

There is a function related to `sort()` called `order()`, which gives us the order of indices.
```{r}
order(x)

x[ order(x) ]
```

> Q13. What are the five least liked candy types in this set?

```{r}
ord <- order(candy$winpercent)
head(candy[ord,])
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
ord <- order(candy$winpercent, decreasing = TRUE)
head(candy[ord,])
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, reorder( rownames(candy), winpercent)) +
  geom_col()
```
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```
> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruit candy?

Starburst

## Taking a look at pricepercent

```{r}
my_cols[as.logical(candy$fruity)]
```

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols)
```
### Add some labels

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label = rownames(candy)) +
  geom_point(col = my_cols) +
  geom_text()
```

```{r}
#install.packages("ggrepel")
library(ggrepel)

# This will give us a cleaner graph

# How about a plot of pricepercent vs winpercent
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

## Exploring the correlation structure

Pearson correlation goes between -1 and +1 with zero indicating no correlation and values close to one being very highly correlated. 

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot, what two variables are anti-correlated?

Chocolate and fruit and anti-correlated.

> Q23. What two variables are most positively correlated?

Chocolate and winpercent or chocolate and bar are the most positively correlated.


# 6. Principal Component Analysis

The base R function for PCA is called `prcomp()` and we can set "scale=TRUE/FALSE".

```{r}
pca <- prcomp(candy, scale = TRUE)

summary(pca)
```

The main result of PCA - i.e. the new PC plot (projection of candy on our new PC axis) is contained in `pca$x`.

```{r}
pc <- as.data.frame(pca$x)

ggplot(pc) + 
  aes(PC1, PC2, label = rownames(pc)) +
  geom_point(col = my_cols)
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus

```{r}
par(mar=c(8, 4, 2, 2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

