---
title: "Lab 12: RNASeq Analysis"
author: "Alex Cagle"
date: "2023-05-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("BiocManager")
#BiocManager::install()
```

```{r}
#BiocManager::install("DESeq2")
```

```{r}
library(BiocManager)
library(DESeq2)
```

In this lab, we will use the DESeq2 package for RNASeq analysis. The data comes from a study of smooth muscle cells in our airway treated with dexamethasone, a glucocorticoid steroid.

## Import Data

```{r}
# Importing countData and colData
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

```{r}
head(counts)
```

```{r}
head(metadata)
```

> Q1. How many genes are in this dataset?

```{r}
nrow(counts)
```
> Q2. How many ‘control’ cell lines do we have?

```{r}
table(metadata$dex)

sum(metadata$dex == 'control')
```

```{r}
# Find which columns in countData are control
# First, look at metadata ($dex) to get indices
control <- metadata[metadata[,"dex"] == "control",]
control
```


```{r}
# Extract all control columns from countData
control.counts <- counts[ ,control$id]
control.counts
```

```{r}
# Calculate the mean value across the rows of control.counts (for each gene in the control sample)
control.mean <- rowMeans(control.counts, na.rm = TRUE)
head(control.mean)

#control.mean <- rowSums( control.counts )/4 
```
> Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r}
# Now, do the same for treated
treated <- metadata[metadata[,"dex"] == "treated",]

treated.counts <- counts[ ,treated$id]

treated.mean <- rowMeans(treated.counts, na.rm = TRUE)
head(treated.mean)
```

```{r}
# Combine means into new dataframe
meancounts <- data.frame(control.mean, treated.mean)
head(meancounts)
```

```{r}
colSums(meancounts)
```
> Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

```{r}
plot(meancounts)
```

> Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this? 

```{r}
plot(meancounts, log = "xy")
```

```{r}
# We can use log transforms for skewed data such as this and because we really care most about relative changes in magnitude.
log2(20/20)
log2(10/20)
log2(40/20)
log2(1.5)
```
```{r}
meancounts$log2fc <- log2(meancounts$treated.mean / meancounts$control.mean)
head(meancounts)
```

> Q. How many genes are up-regulated at the common threshold of +2 log2FC values?

```{r}
sum(meancounts$log2fc >= 2, na.rm = TRUE)
```
What about the stats? These are big changes, but are t hey significant? To do this, we will turn to DESeq2.

## DESeq2 analysis

```{r}
library(DESeq2)
```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

```{r}
# To run main analysis:
dds <- DESeq(dds)
```

```{r}
# To get the results:
res <- results(dds)
head(res)
```

```{r}
plot(res$log2FoldChange, res$padj)
```
It is the low p-value that we care about, which are lost in the skewed plot above. So, let's perform a log transform on the padj values.

```{r}
plot(res$log2FoldChange, -log(res$padj))
abline(v = c(+2, -2), col = 'blue')
abline(h = -log(0.05), col = 'blue')

```

```{r}
# Finally, we can make a color vector to use in the plot to better highlight the genes we care about.
mycols <- rep("grey", nrow(res))
mycols[abs(res$log2FoldChange) >= 2] <- 'red'
mycols[res$padj > 0.05] <- 'grey'

plot(res$log2FoldChange, -log(res$padj), col=mycols)
abline(v = c(+2, -2), col = 'green')
abline(h = -log(0.05), col = 'green')
```

