---
title: "Class 06: R Functions"
author: "Alex Cagle"
date: "2023-04-25"
output: pdf_document
---

# Lab 6

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Loading data
gradebook <- read.csv(url('https://tinyurl.com/gradeinput'), row.names = 1)

head(gradebook)
```

```{r}
# Example input vectors to start with
# student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
# student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
# student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

```{r}
# mean(student1)

# mean(student2, na.rm=TRUE)
```

```{r}
# student2[is.na(student2)] <- 0
# student2
```

```{r}
# x <- student1
# x
# which.min(x)
```

> Q1. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: "[https://tinyurl.com/gradeinput"](https://tinyurl.com/gradeinput”) [3pts]

```{r}
# Question 1:
# Function to calculate average grade with lowest score dropped
grade <- function(x) {
  x[is.na(x)] <- 0
  
  print(x)
  
  lowest_index <- which.min(x)
  
  x <- x[-lowest_index]
  
  mean(x)
  
  # avg <- mean(new_gradebook)
  
  # return(avg)
}
```

```{r}
# Testing grade() with example vectors
# grade(student1)

# grade(student2)

# grade(student3)
```

```{r}
# Question 1:
# Applying grade() function to gradebook
avgs <- apply(gradebook, 1, grade)

# Displaying the average homework grade for each student
avg_df <- data.frame(avgs)
avg_df
```

> Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]

```{r}
# Question 2:
# Highest scoring student
print("The highest score is:")
highest_score <- avg_df[which.max(avg_df[,1]),]
highest_score

print("The highest scoring student is:")
rownames(avg_df)[avg_df[, 1] == highest_score]
```

> Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

```{r}
# Question 3:
# Toughest homework
avg_hw_score <- apply(gradebook, 2, sum, na.rm=T)
avg_hw_score

print("The toughest homework is:")
avg_hw_score[which.min(avg_hw_score)]
```

> Q4. Optional Extension: From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

```{r}
hw <- gradebook
hw

hw$hw1
```

```{r}
cor(hw$hw1, avgs)
cor(hw$hw3, avgs)
```

```{r}
# Masking all NA values to zero
mask <- hw
mask[is.na(mask)] <- 0
mask
```

```{r}
cor(mask$hw5, avgs)
```

```{r}
apply(mask, 2, cor, y=avgs)
```
